Implementing Distributed Systems Directly\\with I/O Automata in ioa++

### 2011-10-01 papers/paper1/case_study.tex
revised motivation at the start of the case study

In this section, we demonstrate that ioa++ can be used to build real distributed systems.
Real distributed systems are designed using a layered approach where the upper layers contain semantically rich business logic while the bottom layers contain low-level network message-passing.
The middle layers, i.e., middleware, contain generic high-level coordination mechanisms that abstract away the low-level details of message-passing.
The lower layers contain abstractions of operating system services such as TCP/UDP sockets (for communicating with other peers) and timers (for detecting time-outs and performing periodic tasks), e.g., \cite{schmidt2001c++},~\cite{schmidt2002c++}.
We apply this layered architecture technique to the simple but representative network programming problem of electing a leader in a unidirectional ring of processes connected by TCP sockets.

The ``business logic'' we will implement is quite simple and consists of periodically circulating a token around a ring of nodes.
If a node has not received a token in a specified amount of time, it initiates a leader election protocol.
Once a leader has been elected, the leader injects a token and the protocol resumes.
Nodes can join and leave the ring at any time.
The protocol should eventually succeed for complete rings.
These dynamics admit fault situations where a ring might lose its leader or contain multiple leaders.
A ring containing multiple leaders should stabilize to having a single leader.
A rare (but possible and valid) situation is a ring circulating a token without a leader, i.e., the leader sends a token, leaves the ring, and is replaced before the token makes a full cycle.
This situation requires no action as a token is circulating in the ring.
We divide the problem into two sub-problems:  leader election and ring formation.

@@

The tcp\_lcr\_automaton which is listed in examples/tcp\_lcr.cpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}} of the ioa++ package implements the complete protocol by composing an asynch\_lcr\_automaton and a tcp\_ring\_automaton.
Figure~\ref{tcp_lcr_automaton} shows the layered architecture of the TCP LCR system.
The bottom layer consists of component automata for TCP sockets, the middle layer consists of a component automaton that supports coordination via a unidirectional ring, and the top layer implements the business logic.
Thus, ioa++ supports features we expect from a component framework for distributed systems.

>>>>

In this section, we illustrate how ioa++ can be used to build real distributed systems.
In practice, distributed systems are designed using a layered approach where the highest layers contain (often semantically rich) application logic.
The lower layers provide services such as TCP/UDP sockets (for communicating between endsystems) and timers (for detecting time-outs and performing periodic tasks), e.g., \cite{schmidt2001c++},~\cite{schmidt2002c++}.
The intermediate layers, i.e., frameworks like ioa++, contain generic coordination mechanisms that serve to abstract away the low-level details from the application logic.

We apply this layered architecture technique to the simple but representative network programming problem of electing a leader in a unidirectional ring of processes connected by TCP sockets.  
In order to focus on how ioa++ supports distributed applications in general, rather than on the details of any
particular application, the application logic in this example consists simply of periodically circulating a token around a ring of nodes.
If a node has not received a token in a specified amount of time it initiates a leader election protocol.
Once a leader has been elected, the leader injects a token and the protocol resumes.

Despite the simplicity of the application logic, the resulting \emph{dynamic} distributed system serves to motivate why and illustrate how a number of crucial features are provided by ioa++.
For example, nodes can join and leave the ring at any time, and yet the protocol should eventually succeed for complete rings when they stabilize.
These dynamics admit fault situations where a ring might lose its leader or temporarily contain multiple leaders, and yet each ring should stabilize to having a single leader.
Another possible and valid (if potentially rare) situation is a ring circulating a token without a leader, i.e., the leader sends a token, leaves the ring, and is replaced before the token makes a full cycle.
% This situation requires no action as a token is circulating in the ring.
We divide these problems into two sub-problems:  leader election and ring formation.

@@

The tcp\_lcr\_automaton which is listed in examples/tcp\_lcr.cpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}} of the ioa++ package implements the complete protocol by composing an asynch\_lcr\_automaton and a tcp\_ring\_automaton.

Figure~\ref{tcp_lcr_automaton} shows the resulting layered architecture of the TCP LCR system.
The bottom layer consists of component automata for TCP sockets, the middle layer consists of a component automaton that supports coordination via a unidirectional ring, and the top layer implements the application logic.
As this simple but representative example illustrates, the ioa++ framework readily supports the development of architecturally layered component-based distributed systems involving non-trivial dynamics.
